---
title: "Prediction of Exercise Quality"
author: "F. Alex Crofut"
date: "Sunday, May 22, 2016"
output: html_document
---

#Executive Summary



#Getting Data and Preprocessing
<<<<<<< HEAD
```{r set seed}
seedme <- 123
set.seed(seedme)
```
For reproducibility, the seed is set to `r format(seedme, scientific=FALSE)`. 

###Required Packages
```{r packages, echo=TRUE, warning=FALSE, message=FALSE}
=======

###Required Packages
```{r packages, echo=FALSE, warning=FALSE, message=FALSE}
>>>>>>> 8b7e6f0e96dee1fbe2224414c8a3629247144cd4
if (!"caret" %in% installed.packages()) install.packages("caret", repos="http://cran.rstudio.com/")
library("caret")
if (!"randomForest" %in% installed.packages()) install.packages("randomForest", repos="http://cran.rstudio.com/")
library("randomForest")
if (!"RCurl" %in% installed.packages()) install.packages("RCurl", repos="http://cran.rstudio.com/")
library("RCurl")
if (!"rpart" %in% installed.packages()) install.packages("rpart", repos="http://cran.rstudio.com/")
library("rpart")
<<<<<<< HEAD
if (!"rpart.plot" %in% installed.packages()) install.packages("rpart.plot", repos="http://cran.rstudio.com/")
library("rpart.plot")
=======
>>>>>>> 8b7e6f0e96dee1fbe2224414c8a3629247144cd4
if (!"rattle" %in% installed.packages()) install.packages("rattle", repos="http://cran.rstudio.com/")
library("rattle")
```

###Download Data
Exercise data was downloaded from the following URLs:

* [Training Data](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv)
* [Test Data](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv)

The data is in the form of two comma-delimited files, one for training data and one for testing data.  The data have many missing data points represented by "NA", "DIV/0" (a divide by zero error), and empty strings.  While reading the data, these strings are all converted to the standard "NA". 
<<<<<<< HEAD
```{r getdata}
=======
```{r getdata, cache=TRUE}
>>>>>>> 8b7e6f0e96dee1fbe2224414c8a3629247144cd4
urlTrain <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
urlTest <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
fileTrain <- getURL(urlTrain)
fileTest <- getURL(urlTest)
datTrain <- read.csv(textConnection(fileTrain), header=TRUE, sep=",", na.strings=c("NA", "#DIV/0!", ""))
datTest <- read.csv(textConnection(fileTest), header=TRUE, sep=",", na.strings=c("NA", "#DIV/0!", ""))
```

###Clean the Data
The training data is comprised of `r format(dim(datTrain)[1], big.mark=",")` observations of `r dim(datTrain)[2]` variables. The variables can be summarized thusly:

1. The first column is a row index
2. Column 2 contains the name of the person performing the exercise
3. Columns 3-7 hold time-related variables related to each observation of the sensor readings
4. Columns 8-159 contain the sensor reading variables, helpfully identified by a string indicated the exercise performed
5. Column 160 contains the "classe" variable that is to be predicted

```{r cleaning, echo=FALSE}
# cast data as numeric
for(i in c(8:159)) {datTrain[,i] <- as.numeric(as.character(datTrain[,i]))}
for(i in c(8:159)) {datTest[,i] <- as.numeric(as.character(datTest[,i]))}

# remove values with too many NAs
threshna <- 0.75
countna <- is.na(datTrain)
dropcols <- which(colSums(countna) > threshna * dim(datTrain))
datTrain <- datTrain[, -dropcols]
<<<<<<< HEAD
datTest <- datTest[, -dropcols]

# remove first column
datTrain <- datTrain[, -1]
datTest <- datTest[, -1]
```
Everything from column 8 through 159 are recast as numeric as they hold data from the sensors.  Many variables are comprised mostly of missing values.  This analysis uses a threshold of `r threshna * 100`% missing data to determine whether or not a variable will be included.  The first column holds a record number and is not useful for prediction.  It is therefore removed as well.  A quick exploration of the data shows that there does appear to be a relationship between time and performance.  Given the time frame, this may be related to tiring as exercises are performed.  Therefore, the date columns were retained.  This leaves `r dim(datTrain)[2] - 1` variables to use as predictors.  

#Training the Predictor
First, the data is separated into a training sample and a quiz sample so an estimate of out-of-sample error can be made before predicting with the test data.
```{r create partitions}
inTrain <- createDataPartition(y=datTrain$classe, p=0.75, list=FALSE)
myTraining <- datTrain[inTrain,]
myQuiz <- datTrain[-inTrain,]
```

###Decision Tree
```{r model decision tree}
modDecTree <- rpart(classe ~ ., data=myTraining, method="class")         # train the model
predDecTree <- predict(modDecTree, myTraining, type="class")            # predict on training set
resultDecTree <- confusionMatrix(predDecTree, myTraining$classe)        # evaluate model success
as.table(resultDecTree$overall)
```
The decision tree yielded an accuracy of `r round(resultDecTree$overall["Accuracy"] * 100, digits=1)`%, meaning an in-sample error rate of `r 100 - round(resultDecTree$overall["Accuracy"] * 100, digits=1)`%.

###Random Forest
```{r model random forest, cache=TRUE}
modRandForest <- randomForest(classe ~ ., data=myTraining)
predRandForest <- predict(modRandForest, myTraining, type="class")
resultRandForest <- confusionMatrix(predRandForest, myTraining$classe)
as.table(resultRandForest$overall)
```
The decision tree yielded an accuracy of `r round(resultRandForest$overall["Accuracy"] * 100, digits=1)`%, meaning an in-sample error rate of `r 100 - round(resultRandForest$overall["Accuracy"] * 100, digits=1)`%. This suggests the random forest model may be overfitted.

###Estimating Out-of-Sample Error
The random forest had the higher accuracy and is therefore used to predict on the test data and estimate out-of-sample error.  The random forest requires coercion of the data in the test sample to the same type as the data in the training sample.
```{r out of sample error}
# predict on quiz sample
predQuiz <- predict(modRandForest, newdata=myQuiz, type="class")
resultQuiz <- confusionMatrix(predQuiz, myQuiz$classe)
as.table(resultQuiz$overall)
```
With an accuracy of `r round(resultQuiz$overall["Accuracy"] * 100, digits=1)`%, the out-of-sample error rate is estimated as `r 100 - round(resultQuiz$overall["Accuracy"] * 100, digits=1)`%

#Conclusion
With an out-of-sample error well above 99%, the model is used to predict classe values on the test data provided.
```{r test data}
# make types in datTest match datTrain
for (i in 1:length(datTest) ) {
        levels(datTest[,i]) <- levels(myTraining[,i])     
}

predTest <- predict(modRandForest, newdata=datTest, type="class")
predictions <- data.frame(index=c(1:20), classe=predTest)
predictions
```
=======

# remove first column
datTrain <- datTrain[, -1]
```

Everything from column 8 through 159 are recast as numeric as they hold data from the sensors.  Many variables are comprised mostly of missing values.  This analysis uses a threshold of `r threshna * 100`% missing data to determine whether or not a variable will be included.  The first column holds a record number and is not useful for prediction.  It is therefore removed as well.  A quick exploration of the data shows that there does appear to be a relationship between time and performance.  Given the time frame, this may be related to tiring as exercises are performed.  The time measures are therefore left in the analysis for further evaluation.  This leaves `r dim(datTrain)[2]` variables to use as predictors.  

#Training the Predictor

```{r prep, echo=FALSE}
seedme <- 25897
set.seed(seedme)
```
To insure reproducibility, the seed is set to `r seedme`.  

###Decision Tree
```{r model decision tree}
mdlDecTree <- rpart(classe ~ ., data=datTrain, method="class")
fancyRpartPlot(mdlDecTree)
```

###Random Forest
```{r model random forest}


```

###PCA
```{r model PCA}


```


#Predicting on the Test Data


#Conclusion
>>>>>>> 8b7e6f0e96dee1fbe2224414c8a3629247144cd4
